### 2022-09-15 reports/paper/01-introduction.rmd
Another draft of  manuscript

Metalworking fluids (MWF) are complex chemical mixtures used to lubricate and cool machinery during manufacturing processes. Straight MWF are the oldest class of MWF, comprising of simple oils or combinations of oils. By the mid 20th century, straight MWF waned in popularity and ceded market share to soluble MWF. Soluble MWF are water-oil mixtures with diverse additives for improving performance and durability. By the 1970s, semisynthetic and synthetic MWF, which contain little to no oil, overtook straight oils in market share. Nonetheless, at the turn of the millennium, soluble MWFs still retained the largest share of the MWF market [@Childers_2005]. Carcinogenicity of MWFs has been of concern since as early as the 1970s, when the International Agency for Research on Cancer (IARC) classified mineral oils as carcinogenic, citing studies of occupational exposures among workers in oil, textile, and metal industries [@IARC_1973]. Concerns over the health effects of straight MWF hastened the adoption of soluble MWF, which contain less oil per volume, but a richer cocktail of additives including chromates, cycloalkanes, phenols, organochlorines, nitrosamines, and triazines: classes of chemicals with known or plausible carcinogenicity [@IARC_1987;@IARC_2014;@IARC_2017]. Health and environmental considerations led to substantial variation in MWF composition over time, but the elimination of particular etiologic agents often leads to their replacement with other potentially hazardous chemical compounds. Exposure limits for MWF as a mixture encourage limiting occupational exposure to multiple agents simultaneously.

Here, we assess the effect of hypothetical exposure limits on non-Hodgkin lymphoma (NHL) cumulative incidence (risk) from 1985 to 2005 in the United Auto Workers-General Motors (UAW-GM) Occupational Cohort Study. The large size of the study population and rich time-varying, quantitative MWF exposure data provide an opportunity to study this relatively rare cancer while contrasting realistic interventions on MWF exposure in a longitudinal cohort setting. Though NHL accounts for a small proportion of cancer incidence of all sites, it has garnered intense research attention because its incidence nearly quadrupled since 1960 [@Ekstrom-Smedby_2006]. Immune modulation due to immunosuppressive therapy and infection by human immunodeficiency virus (HIV) are the most prominent risk factors of NHL, but they cannot fully explain this rise in NHL incidence [@Shiels_2013;@Chiu_2015]. Epidemiologic investigations of environmental and occupational exposures have yielded varied results [@Ekstrom-Smedby_2006;@Alexander_2007]. This variation in epidemiologic findings may reflect the substantial etiologic heterogeneity of NHL, but studies of specific NHL subtypes are limited by the lack of quantitative exposure information and the lack of control for the healthy worker survivor effect (HWSE) or other forms of time-varying confounding affected by prior exposure [@Evens_2008;@Morton_2014]. The HWSE, a key concern in occupational epidemiology, is the dynamic selection process by which healthier individuals remain at work, where they accumulate more occupational exposure, and less healthy individuals leave work [@Arrighi_1994]. Standard analyses of occupational cohorts affected by the HWSE result in a downward bias on the exposure-outcome associations of interest.

In the present analysis of the UAW-GM Cohort Study, we estimated NHL risk from 1985 to 2005 under hypothetical limits on average annual soluble MWF by applying the novel hazard-extended iterative conditional expectation (ICE) parametric g-formula estimator [@Wen_2020]. We leveraged time-varying quantitative MWF exposure data in tandem with employment data to adjust for the HWSE. In particular, we estimated the expected number of NHL cases we would observe if the hypothetical limits of 0.5, 0.25, and 0.05 mg/m^3^ were enforced for soluble MWF over workers' entire working lifetimes. Since workers were exposed to MWF of other types as well, we also evaluated the effect of dynamically reducing soluble MWF exposure with the goal of limiting total MWF exposure to 0.5 and 0.25 mg/m^3^.

>>>>

Non-Hodgkin Lymphoma (NHL) is the seventh most common cancer type in the United States and has garnered intense research attention because its incidence in the United States doubled between 1973 and 1994, before plateauing at around 19 per 100,000 persons [@seer_1994;@Ekstrom-Smedby_2006]. The strongest known risk factor of NHL is immunosuppression, both congenital and acquired [@Filipovich_1992;@Chiu_2015]. Modulation of immune function contributes to the pathogenesis of NHL directly by perturbing lymphocyte function or indirectly by reducing immunosurveillance against neoplasms. Immunodeficiency may also interact with infection with Epstein-Barr Virus, resulting in uncontrolled B-cell proliferation [@Oertel_2002]. Acquired immunodeficiency through infection with human immunodeficiency virus (HIV) can lead to genetic lesions and dysregulation while increasing vulnerability to infection by other oncogenic agents [@Grulich_1999;@Alexander_2007]. However, since immunodeficiency and infection with HIV are rare, they cannot fully explain the historic rise or present burden of NHL [@Shiels_2013]. Since the rise in NHL incidence coincided with a period of rapid and extensive chemicalization in industry, agriculture and warfare, there has been stronger interest in the environmental and occupational epidemiology on NHL [@Nelson_2005;@Romero_2021]. A meta-analysis of 44 articles published between 1980 and 2014 found statistically significant associations between NHL and exposure to pesticides of several classes including carbamate, organophosporus, triazine, and organochlorine [@Schinasi_2014]. While this meta-analysis presents striking evidence of the potential explanatory role of occupational exposures, it also highlights the limited breadth of occupational exposures studied.

Studies of NHL and other important occupational exposures including oils, solvents, and lubricants remain inconclusive [@Ekstrom-Smedby_2006]. Furthermore, since NHL is relatively rare, nearly all studies of NHL incidence follow a single time-point case-control design [@Schinasi_2014;@Alexander_2007]. Large multicenter studies including EPILYMPH and InterLymph attained large enough numbers of cases to investigate particular NHL subtypes, but despite their impressive size and comprehensiveness, were limited by the lack of quantitative time-varying exposure information and the lack of adjustment for time-varying sources of bias [@Besson_2006;@Morton_2014]. In the case of occupational studies, the healthy worker survivor effect (HWSE) is a key concern. The HWSE is the dynamic selection process by which healthier individuals remain at work, where they accumulate more occupational exposure, and less healthy individuals leave work [@Arrighi_1994]. Standard analyses of occupational cohorts affected by the HWSE result in a downward bias on the exposure-outcome associations of interest, as less healthy workers typically accrue less exposure than healthier workers.

Metalworking fluids (MWF), an important class of occupational exposures, are complex mixtures of oils and chemicals used to lubricate and cool metal during manufacturing processes. During use, MWFs aerosolize and present a potential health hazard to exposed workers. The type of MWF applied depends in part on the particular metalworking operation. Straight MWF, the oldest class of MWF, is comprised mostly of mineral oil. Concerns over the carcinogenicity of MWF began as early as the 1970s, when the International Agency for Research on Cancer (IARC) classified mineral oils as carcinogenic, citing studies of occupational exposures among workers in oil, textile, and metal industries [@IARC_1973]. The combination of health concerns, the high cost of oil in the 1970s, and the introduction of performance-enhancing water-soluble chemical additives led to the gradual replacement of straight MWF with water-based soluble and synthetic MWFs for many, but not all operations [@Childers_2005]. The elimination of particular etiologic agents often leads to their replacement with other potentially hazardous chemical compounds, however [@Kassotis_2020;@Maertens_2021]. Although water-based MWF contain little to no mineral oil, they contain a richer cocktail of additives including chromates, cycloalkanes, phenols, organochlorines, nitrites, sulfonates, and triazines: classes of chemicals with known or plausible carcinogenicity [@IARC_1987;@IARC_2014;@IARC_2017]. Hence, limits on entire classes of exposures grouped by functional use may result in stronger, more lasting protections for worker health than limits on particular etiologic agents because they will remain in place even as the composition and formulation of complex occupational exposures change [@Maertens_2021]. More than 20 years after the IARC classification of mineral oils as carcinogenic, the National Institute for Occupational Safety and Health (NIOSH) concluded that there exists substantial evidence linking all MWF exposure to several different cancers and now recommends an exposure limit of 0.5 mg/m^3^ for total particulate matter derived from any type of MWF [@niosh_1998;@Mirer_2003].

Here, we assess the effect of hypothetical MWF exposure limits on NHL cumulative incidence in the United Auto Workers-General Motors (UAW-GM) Occupational Cohort Study. In particular, we assess the effect of limits on soluble MWF that (1) ignore exposure to other types of MWF and (2) account for other types of MWF exposure by intervening more strongly on soluble MWF when exposure to all types of MWF is high. The latter approach is more consistent with the NIOSH REL, which was recommended for exposure to any type of MWF. The large size of the study population and rich time-varying, quantitative MWF exposure data provide an opportunity to study this relatively rare cancer and evaluate realistic interventions on MWF exposure in a longitudinal cohort setting. A recent study of cancer incidence in this iconic occupational cohort found a monotonic association between NHL and exposure to soluble MWF [@Colbeth_2022]. This association was based on a traditional Cox proportional hazards regression, a method vulnerable to bias due to confounding by time-varying factors affected by prior exposure. We therefore explore this association using a novel causal estimator capable of adjusting for time-varying confounding and censoring affected by past exposure. This estimator also yields marginal effect estimates of hypothetical interventions, which are more compatible with a population health framework than conditional measures.

### 2022-09-15 reports/paper/02-methods.rmd
Another draft of  manuscript

@@

The UAW-GM Cohort includes hourly workers at three automobile manufacturing plants in Michigan who had worked at least 3 years by 1985. Past papers provide detailed descriptions of the cohort [@Eisen_1992;@Eisen_2001]. The present study population (N = 34,738) was restricted to the autoworkers who were at work in 1941 or not yet hired, alive at the start of follow-up, and missing no more than half of their employment history. Autoworkers in the study population were followed for NHL incidence from January 1, 1995 until NHL diagnosis, death, December 31, 2004, or upon reaching the oldest observed age at death, whichever came earlier.

@@

We identified incident cancers in the UAW-GM Cohort that occurred between 1985 and 2004 by linkage to the Michigan Cancer Registry (MCR). Workers at Plants 1 and 2, located in the greater Detroit metropolitan area, were also linked to the Detroit Regional Registry of the Surveillance, Epidemiology, and End Results (SEER) Program. Cancer types were distinguished using site and histology codes conforming to the International classification of Diseases for Oncology, 3^rd^ edition (ICD-O-3). Non-Hodgkin lymphoma was defined by cancers with any of the following ICD-O-3 Histology codes: 9590-9597, 9670-9671, 9673, 9675, 9678-9680, 9684, 9687-9691, 9695, 9698-9702, 9705, 9708-9709, 9712, 9714-9719, 9724-9729, 9735, 9737-9738, 9811-9818, 9823, 9827, 9837. Details regarding cancer incidence follow-up are described elsewhere [@Colbeth_2022]. Vital status was ascertained by company records and by linkage to Social Security Administration, National Death Index, death certificate, and state mortality files.

Covariates including year of hire, sex, race, and plant location were obtained from company records. Race data were missing for about 16% of the workers. In analyses, missing race was considered a distinct category of race.

@@

For each combination of job type, department, and plant industrial hygienists collected several hundred personal and area samples for particulate matter (mg/m^3^) over the course of three decades to derive quantitative 8-hour time-weighted average exposure estimates to soluble, straight, and synthetic MWFs. Workers' time-weighted average annual exposure to each MWF type was determined according to employment records, which recorded time-varying job type, department, and plant. For employment records that were no less than half complete, gaps in the record were interpolated by carrying forward the last known job type. In each year, cumulative exposures to soluble, straight, and synthetic MWFs (mg/m^3^-years) were calculated by taking cumulative sums of average annual exposures. In analyses, MWF exposure was lagged 10 years. Exposure assessment is described in detail elsewhere [@Hallock_1994;@Woskie_1994;@Woskie_2003].

@@

We applied the hazard-extended ICE parametric g-formula with pooling over treatment history [@Wen_2020] to estimate 20-year counterfactual risk of non-Hodgkin lymphoma under five hypothetical stochastic interventions on exposure to soluble MWFs with the elimination of censoring/competing risks. The first three interventions were static interventions that limited average annual exposure to soluble MWF to 0.5, 0.25, and 0.05 mg/m^3^. The remaining two were dynamic interventions that reduced average annual exposure to soluble MWF so that exposure to all MWF types was capped at 0.5 and 0.25 mg/m^3^, if possible. If those limits on total average annual exposure were not possible by intervention on soluble MWF alone, average annual exposure to soluble MWF was reduced to 0. Interventions were applied at hire, before the start of follow-up, through the end of follow-up. Person-years in which average annual exposure was below the hypothetical limits were not intervened upon. These hypothetical exposure limits reflect the Recommended Exposure Limit (REL) of the National Institute for Occupational Safety and Health (NIOSH) for total particulate mass from MWFs (0.5 mg/m^3^), half the REL (0.25 mg/m^3^), and a tenth of the REL (0.05 mg/m^3^) [@niosh_1998].

The hazard-extended ICE parametric g-formula estimator is an ICE expression of the parametric g-formula estimator, further extended to incorporate the predicted hazard of past outcomes. This differs from the classical ICE estimator, which incorporates past outcome data as indicators [@Wen_2020]. Under the assumptions of conditional exchangeability at all time points, positivity, counterfactual consistency, and correct model specification, the hazard-extended parametric g-formula yields unbiased estimates of counterfactual risk with greater statistical efficiency than both propensity-score based estimators and the classical ICE g-formula.

Our implementation of the hazard-extended ICE parametric g-formula involved a series of model-based standardization steps using logistic regression. Post-intervention estimates of the discrete hazard of NHL given all exposures and covariates were combined iteratively from the end of follow-up to the start. In each iteration, predicted discrete hazards were standardized over post-intervention exposure and covariate histories before combining with discrete hazards from the previous iteration. The result of this iterative process was sequentially-standardized estimates of the NHL risk over the entire follow-up period. Averaging over the baseline distribution of covariates yielded the counterfactual risk estimate of NHL when the intervention of interest was enforced for the entire study population.

We split the 20-year follow-up period into eight periods; the first two periods spanned four years each, and the remaining six periods spanned two years each. Post-intervention exposure and exposure history were summarized as cumulative exposure. We modeled discrete hazards by fitting a pooled logistic regression for NHL over at-risk person-periods given cumulative exposure to straight, soluble, and synthetic MWFs, employment status, cumulative time off, year of hire, sex (male/female), race (Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3). Cumulative exposure to MWFs, employment status, and cumulative time off were lagged 10 years. All continuous variables were represented as categorical variables with cut points determined by the tertiles of nonzero values among NHL cases. During the iterative combination of discrete hazards, we performed model-based standardization over baseline covariates and the complete set time-varying covariate histories.

We estimated risk under the observed distribution of soluble MWF exposure (natural course) and under the five interventions. We contrasted the risk under intervention to that under the natural course by computing relative risks. Confidence intervals were computed using the nonparametric bootstrap with 1000 samples and centering on the estimate computed from observed data. In secondary analyses (Appendix), we repeated the analysis to evaluate interventions on exposure to straight and synthetic MWFs.

>>>>

We estimated NHL risk from 1985 to 2005 under hypothetical limits on average annual exposure to soluble MWF by applying the novel hazard-extended iterative conditional expectation (ICE) parametric g-formula estimator [@Wen_2020]. We leveraged time-varying quantitative MWF exposure data in tandem with employment records to adjust for the HWSE. First, we estimated the expected number of NHL cases per 1000 workers that we would observe if there were no competing risks. We contrasted  this counterfactual risk to that when in addition, the hypothetical exposure limits of 0.5, 0.25, and 0.05 mg/m^3^ were enforced for soluble MWF over workers' entire working lifetimes. These hypothetical limits are static interventions because they do not consider other covariates or exposures. However, since workers were often exposed to multiple types of MWF, both daily and cumulatively, we also evaluated the effect of limiting total MWF exposure to those hypothetical limits by intervening only on exposure to soluble MWF. These interventions, which take into consideration exposure to MWF of other types, are dynamic interventions on exposure to soluble MWF. All of the hypothetical interventions we evaluate are stochastic interventions rather than deterministic interventions because post-intervention exposures take on a range of values.

@@

The UAW-GM cohort includes all hourly workers at three automobile manufacturing plants in Michigan who had worked at least 3 years by 1985. Past papers provide detailed descriptions of the cohort [@Eisen_1992;@Eisen_2001]. The present study population (N = 34,738) was restricted to the autoworkers who were at work in 1941 or not yet hired, missing no more than half of their employment history, and still alive at the start of follow-up. Autoworkers in the study population were followed for NHL incidence from January 1, 1985 until NHL diagnosis, death, 10 years after the end of employment records, or upon reaching the oldest observed age at death (108 years), whichever came earlier.

@@

We identified incident cancers in the UAW-GM Cohort that occurred between 1985 and 2004 by linkage to the Michigan Cancer Registry (MCR). Workers at Plants 1 and 2, located in the greater Detroit metropolitan area, were also linked to the Detroit Regional Registry of the Surveillance, Epidemiology, and End Results (SEER) Program. Cancer types were distinguished using site and histology codes conforming to the International classification of Diseases for Oncology, 3^rd^ edition (ICD-O-3). Non-Hodgkin lymphoma was defined by cancers with any of the following ICD-O-3 Histology codes: 9590-9597, 9670-9671, 9673, 9675, 9678-9680, 9684, 9687-9691, 9695, 9698-9702, 9705, 9708-9709, 9712, 9714-9719, 9724-9729, 9735, 9737-9738, 9811-9818, 9823, 9827, 9837. Details regarding cancer incidence follow-up are described elsewhere [@Colbeth_2022]. Vital status was ascertained from company records and by linkage to Social Security Administration, National Death Index, and state mortality files.

Covariates including year of hire, sex, race, and plant location were obtained from company records. Race was missing for about 16% of the cohort, most commonly among workers hired before 1960 in Plant 2. In analyses, missing race was considered a distinct category.

@@

Company industrial hygienists collected several hundred personal and area samples for total particulate matter (mg/m^3^) composed of MWF over many decades. Research industrial hygienists collected additional air sampling data when the cohort study was launched in the mid 1980s. These data were combined with the historical data to derive quantitative 8-hour time-weighted average exposure estimates to soluble, straight, and synthetic MWFs for each combination of job, department, and plant over time. Workers' time-weighted average annual exposure to each MWF type was determined by combining this job-exposure matrix with employment records, which recorded time-varying job type, department, and plant. For employment records that were at least half complete, gaps in the record were interpolated by carrying forward the last known job type. The exposure assessment is described in detail elsewhere [@Hallock_1994;@Woskie_1994;@Woskie_2003]. Previous analyses of NHL applied exposure lags of 1 to 20 years; we lagged cumulative MWF exposures by 10 years to account for disease latency [@Smith_2007;@Karipidis_2007;@Zhang_2019].

@@

We applied the hazard-extended ICE parametric g-formula with pooling over treatment history [@Wen_2020] to estimate 20-year counterfactual risk of non-Hodgkin lymphoma. We considered six stochastic interventions on average annual exposure to soluble MWFs after the elimination of censoring/competing risks. We considered hypothetical exposure limits starting with the NIOSH Recommended Exposure Limit (REL) for total particulate mass (PM) from MWFs (0.5 mg/m^3^). Then, we estimated the risk at half the REL (0.25 mg/m^3^), and a tenth of the REL (0.05 mg/m^3^) [@niosh_1998]. The first three interventions were static interventions which enforced these hypothetical limits on average annual exposure to soluble MWF. The remaining three interventions were dynamic interventions that reduced average annual exposure to soluble MWF with the goal of enforcing the hypothetical limits for total exposure to MWF, regardless of type. If those limits on total average annual exposure were not possible by intervention on exposure to soluble MWF alone, average annual exposure to soluble MWF was reduced to 0. Interventions were applied at hire, before the start of follow-up, through the end of follow-up. Person-years were not intervened upon if the observed average annual exposure was below the hypothetical limit. Figure 1 illustrates the static and dynamic interventions for three person-year examples. Panel A shows the observed levels of exposure to soluble MWF and to MWF of other types. Panel B shows the post-intervention levels of exposure under a static intervention capping exposure to soluble MWF at the hypothetical limit (dotted line). Panel C shows the post-intervention levels under a dynamic intervention reducing exposure to soluble MWF with the goal of capping total MWF exposure at the hypothetical limit. Note that in person-year 3, the observed level of exposure to total MWF (and thus, to soluble MWF) is below the hypothetical limit, so the post-intervention exposure is left at the observed level.

![Figure 1. Observed and post-intervention levels of exposure to soluble MWF and MWF of other types in three example person-years. The white boxes represent exposure to soluble MWF, and the shaded boxes represent exposure to MWF of other types. The horizontal dotted line represents the hypothetical exposure limit.](`r here::here("resources/images", "interventions.png")`){width=6.5in}

The hazard-extended ICE parametric g-formula estimator is an ICE expression of the parametric g-formula estimator further extended to incorporate the predicted hazard of past outcomes [@Wen_2020]. The classical ICE g-formula estimator is algebraically equivalent to the noniterative conditional expectation (NICE) g-formula estimator, which typically involves computing covariate-standardized estimates sequentially, starting at the index time, by specifying parametric models for the joint distribution of the confounders, exposures, and outcomes at each time point [@Taubman_2009]. In contrast, ICE g-formula estimators avoid the need for modeling the entire joint distribution of observed variables and only require outcome models for each time point. Counterfactual outcome estimates over the follow-up period are computed from interval-specific conditional estimates by applying the tower rule of expectation. Under the assumptions of conditional exchangeability at all time points, positivity, counterfactual consistency, and correct model specification, the hazard-extended parametric g-formula yields unbiased estimates of counterfactual risk with greater statistical efficiency than both propensity score-based estimators and the classical ICE g-formula.

Our implementation of the hazard-extended ICE parametric g-formula involved a series of model-based standardization steps using logistic regression. Post-intervention estimates of the discrete hazard of NHL given all exposures and covariates were combined iteratively from the end of follow-up to the start. In each iteration, predicted discrete hazards were standardized over post-intervention exposure and covariate histories before combining with discrete hazards from the previous iteration. This iterative process resulted in sequentially standardized estimates of NHL risk over the entire follow-up period. Averaging over the baseline distribution of covariates yielded the counterfactual risk estimate of NHL when the intervention of interest was enforced for the entire study population.

We split the 20-year follow-up period into eight time periods; the first two periods spanned four years each, and the remaining six periods spanned two years each. The first two periods are longer in length to account for the smaller number of cases in those years. Post-intervention exposure and exposure history were summarized as cumulative exposure. We modeled discrete hazards by fitting a pooled logistic regression for NHL over at-risk person-periods given cumulative exposure to straight, soluble, and synthetic MWFs, employment status, cumulative time off, year of hire, sex (male/female), race (Black/white/unknown), and plant (Plant 1/Plant 2/Plant 3). Cumulative exposure to MWFs, employment status, and cumulative time off were lagged 10 years. All continuous variables were represented as categorical variables with cut points determined by the tertiles of nonzero values among NHL cases. During the iterative combination of discrete hazards, we performed model-based standardization over baseline covariates and the complete set of time-varying covariate histories.

We estimated risk under the observed distribution of soluble MWF exposure (natural course) and under the six interventions. We contrasted the risk under intervention to that under the natural course by computing relative risks. Confidence intervals were computed using the nonparametric bootstrap with 1000 samples and centering on the estimate computed from observed data.
